---
title: "Example {tidytof} workflows"
subtitle: "Supplementary Information"
author: "Timothy Keyes, Abhishek Koladiya, Yu Chen Lo, Garry P. Nolan, and Kara L. Davis"
date: "`r Sys.Date()`"
output:
  pdf_document: 
    toc: true
    toc_depth: 2
  officedown::rdocx_document:
    mapstyles:
      Normal: First Paragraph
    reference_docx: reference.docx
    tables: 
      width: 1.0
  html_document:
    df_print: paged
bibliography: supplementary_references.bib
csl: biomed-central.csl
urlcolor: blue
---

```{r setup_workflows, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE 
)

options(tibble.print_min = 4L, tibble.print_max = 4L)

library(tidyverse)
library(tidytof)
library(kableExtra)
library(flextable)
library(officer)
```

In this note, we provide some example code and output demonstrating how to use {tidytof} to perform common CyTOF data analysis tasks. For additional, exhaustive details for each of {tidytof}'s functions, see [the package website](https://keyes-timothy.github.io/tidytof/index.html).

### Accessing the data for this vignette

First, we will download a dataset originally collected for the development of the [CITRUS](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC4084463/) algorithm. These data are available in the `{HDCytoData}` package, which is available on Bioconductor and can be accessed with the following command:

```{r, eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager")
}

BiocManager::install("HDCytoData")
```

To load the CITRUS data into our current R session, we can call a function from the `{HDCytoData}`, which will provide it to us in a format from the `{flowCore}` package (called a "flowSet"). To convert this into a tidy tibble, we can use `{tidytof}` built-in method for converting {flowCore} objects into `tof_tbl`'s .

```{r, message = FALSE, warning = FALSE}
citrus_raw <- HDCytoData::Bodenmiller_BCR_XL_flowSet()

citrus_data <- 
  citrus_raw %>% 
  as_tof_tbl(sep = "_")
```

Thus, we can see that `citrus_data` is a `tof_tbl` with `r nrow(citrus_data)` cells (one in each row) and `r ncol(citrus_data)` pieces of information about each cell (one in each column).

We can also extract some metadata from the raw data and join it with our single-cell data using some functions from the `tidyverse`:

```{r}
citrus_metadata <- 
  tibble(
    file_name = as.character(flowCore::pData(citrus_raw)[[1]]), 
    sample_id = 1:length(file_name),
    patient = stringr::str_extract(file_name, "patient[:digit:]"), 
    stimulation = stringr::str_extract(file_name, "(BCR-XL)|Reference")
  ) %>% 
  mutate(
    stimulation = if_else(stimulation == "Reference", "Basal", stimulation)
  )

citrus_metadata %>%
  head()
```

Thus, we now have sample-level information about which patient each sample was collected from and which stimulation condition ("Basal" or "BCR-XL") each sample was exposed to before data acquisition.

Finally, we can join this metadata with our single-cell `tof_tbl` to obtain the cleaned dataset.

```{r}
citrus_data <- 
  citrus_data %>% 
  left_join(citrus_metadata, by = "sample_id") 
```


After these data cleaning steps, we now have `citrus_data`, a `tof_tbl` containing cells collected from 8 patients. Specifically, 2 samples were taken from each patient: one in which the cells' B-cell receptors were stimulated ("BCR-XL") and one in which they were not ("Basal"). In `citrus_data`, each cell's patient of origin is stored in the `patient` column, and each cell's stimulation condition is stored in the `stimulation` column. In addition, the `population_id` column stores information about cluster labels that were applied to each cell using a combination of FlowSOM clustering and manual merging (for details, run `?HDCytoData::Bodenmiller_BCR_XL` in the R console).

### Analyzing single-cell data

As a `tof_tbl`, `citrus_data` has access to all {dplyr} functions for general data cleaning. For example, we might want to remove some columns that are either redundant or represent measurements that we're not interested in using the `select()` function. In addition, we might want to convert some categorical columns initially represented as numeric vectors into character vectors so that we don't accidentally use them for computation using the `mutate()` function. We can also rename each of our protein measurement columns to omit the metal names using the `rename()` function in order to make column selection a bit easier. 

```{r}
citrus_data <- 
  citrus_data %>% 
  select(
    -Time, 
    -Cell_length,
    -contains("DNA"), 
    -contains("Dd"), 
    -contains("BC", ignore.case = FALSE), 
    -patient_id, 
    -group_id
  ) %>% 
  mutate(
    population_id = as.character(population_id), 
    sample_id = as.character(sample_id)
  ) %>% 
  rename_with(
    .fn = ~ str_remove(.x, pattern = "_.+$"), 
    .cols = c(-file_name, -population_id, -sample_id)
  )
```

We can also apply a common variance-stabilizing transformation (the arcsinh transformation) to each protein measurement column of `citrus_data` using {tidytof}'s `tof_preprocess()` verb: 

```{r}
citrus_data <- 
  citrus_data %>% 
  tof_preprocess()
```

From here, we can try out several workflows for analyzing and visualizing `citrus_data` using various {tidytof} verbs: 


```{r}
citrus_data %>% 
  # cluster cells based on their surface phenotype using flowSOM
  tof_cluster(cluster_cols = starts_with("CD"), method = "flowsom") %>% 
  # downsample cells from each flowsom cluster
  tof_downsample(group_cols = population_id, num_cells = 100, method = "constant") %>% 
  # make a force-directed layout plot 
  tof_plot_cells_layout(knn_cols = starts_with("CD"), color_col = .flowsom_metacluster) 
```


```{r}
citrus_data %>% 
  # downsample within each (pre-identified) cell subpopulation
  tof_downsample(group_cols = population_id, num_cells = 500, method = "constant") %>% 
  # perform dimensionality reduction using UMAP
  tof_reduce_dimensions(method = "umap") %>% 
  # make a plot
  tof_plot_cells_embedding(
    embedding_cols = contains("UMAP"), 
    color_col = population_id
  )
```


```{r}
citrus_data %>% 
  # downsample within each (pre-identified) cell subpopulation
  tof_downsample(group_cols = population_id, num_cells = 500, method = "constant") %>% 
  # perform dimensionality reduction using tsne
  tof_reduce_dimensions(method = "tsne") %>% 
  # make a plot
  tof_plot_cells_embedding(
    embedding_cols = contains("tsne"), 
    color_col = population_id
  )
```

### Analyzing cluster-level data


```{r}
citrus_data %>% 
  # perform differential abundance analysis using paired t-tests with 
  # multiple hypothesis correction
  tof_daa_ttest(
    cluster_col = population_id, 
    group_cols = c(patient, stimulation), 
    effect_col = stimulation, 
    test_type = "paired"
  )
```


```{r}
citrus_data %>% 
  # perform differential expression analysis using paired t-tests with 
  # multiple hypothesis correction
  tof_dea_ttest(
    cluster_col = population_id,
    marker_cols = where(is.double), 
    group_cols = c(patient, stimulation), 
    effect_col = stimulation, 
    test_type = "paired"
  ) %>%
  # select only the significant cluster-marker pairs 
  dplyr::filter(significant == "*")
```


```{r, eval = FALSE}
citrus_data %>% 
  # compute the mean of each marker for each cluster in each patient 
  tof_extract_central_tendency(
    cluster_col = population_id, 
    group_cols = patient, 
    marker_cols = where(is.double),
    central_tendency_function = mean
  ) 
```

```{r}
citrus_data %>% 
  # compute the abundance of each cluster in each  patient
  tof_extract_proportion(
    cluster_col = population_id, 
    group_cols = patient 
  )
```


```{r}
citrus_data %>% 
  # compute the median of each marker for each cluster in each patient 
  tof_extract_central_tendency(
    cluster_col = population_id, 
    group_cols = patient, 
    marker_cols = where(is.double),
    central_tendency_function = median, 
    # request in long format, i.e. suitable for {ggplot2}
    format = "long"
  ) %>% 
  # only look at pSTAT1 
  dplyr::filter(stringr::str_detect(channel, "pStat1")) %>% 
  # make a plot using {ggplot2}
  ggplot(aes(x = population_id, y = values, fill = patient)) + 
  geom_point(shape = 21) + 
  labs(subtitle = "median pSTAT1 expression in each patient") + 
  theme_bw()
```

### Analyzing sample-level data

Using `citrus_data`, we can build a predictive model of for which samples are in the stimulation condition and which ones are in the basal condition solely using signaling phenotypes.  

```{r}
signaling <- 
  c("pNFkB", "pp38", "pStat5", "pAkt", "pStat1", "pStat3", "pBtk", "pS6")

model_result <-
  citrus_data %>% 
  # compute sample-level features from the input single-cell data using only 
  # signaling columns (ones most relevant to the stimulation condition)
  tof_extract_features(
    cluster_col = population_id, 
    group_cols = c(patient, stimulation), 
    lineage_cols = all_of(signaling), 
    signaling_cols = all_of(signaling)
  ) %>% 
  # encode outcome variable as a factor
  mutate(stimulation = as.factor(stimulation)) %>% 
  # split data into 3 cross-validation folds
  tof_split_data(num_cv_folds = 3) %>% 
  # train logistic regression model with lasso regularization
  # using 3-fold cross validation
  tof_train_model(
    predictor_cols = c(-patient, -stimulation), 
    response_col = stimulation, 
    model_type = "two-class", 
    remove_zv_predictors = TRUE, 
    impute_missing_predictors = TRUE,
    hyperparameter_grid = tof_create_grid(
      mixture_values = 1, 
      penalty_values = c(0.001, 0.01, 0.1, 0.4)
    )
  )

# assess the model's performance on the full dataset
model_metrics <- 
  model_result %>% 
  tof_assess_model()


model_result %>% 
  tof_plot_model()
```

